---
title: "Analysis of the Adverse Health and Economic Impacts of US Storms"
author: "Farina Fayyaz"
date: "2024-02-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Synopsis

This research delves into the NOAA Storm Database, spanning 1950 to November 2011, to assess the impact of severe weather events on human health, property, and crops. Through detailed analysis, we aim to identify which types of events pose the greatest threat to   
1. Population  
2. Economy  
measured by injuries, fatalities, and economic consequences.   
For more information about Data: [Documentation][1]

## 2. Data Processing
### 2.1. Downloading the Data

```{r Downloading the Data}
if (!dir.exists("Data")) {
  dir.create("Data")
}
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",destfile = "Data/data.csv")

```

### 2.2 Loading Data into R
```{r Loading Data}
data <- read.csv("Data/data.csv")
```
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
[1]: https://github.com/FarinaFayyaz/datasciencecoursera/tree/main/05.%20Reproducible%20Research/Course%20Project%202