---
title: "Analysis of the Adverse Health and Economic Impacts of US Storms"
author: "Farina Fayyaz"
date: "2024-02-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,echo = TRUE)
```

## 1. Synopsis

This research delves into the NOAA Storm Database, spanning 1950 to November 2011, to assess the impact of severe weather events on human health, property, and crops. Through detailed analysis, we aim to identify which types of events pose the greatest threat to   
1. Population  
2. Economy  
measured by injuries, fatalities, and economic consequences. 
For more information about Data: [Documentation][1]

## 2. Research Questions
1. Across the United States, which types of events are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?

## 3. Data Processing
### 3.1. Downloading the Data

```{r Downloading the Data}
if (!dir.exists("Data")) {
  dir.create("Data")
}
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",destfile = "Data/data.csv")

```

### 3.2 Loading Data into R
```{r Loading Data}
# Loading dependencies
library(data.table)
# Reading Data from .csv file
data <- read.csv("Data/data.csv")
# Converting to data.table
df <- as.data.table(data)
```
### 3.3 Identifying Column Names
```{r Identifying Column Names}
# Identifying Column Names
names(df)
```
### 3.4 Subsetting the Data
We only need to keep columns "EVTYPE", "INJURIES","FATALITIES","PROPDMG","PROPDMGEXP","CROPDMG" and "CROPDMGEXP" for our analysis. Also, we'll only use data where fatalities and injuries occurred.
```{r Subsetting the Data}
# Keep only necessary columns
stormdata <- stormdata[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]



```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
[1]: https://github.com/FarinaFayyaz/datasciencecoursera/tree/main/05.%20Reproducible%20Research/Course%20Project%202